Prelim Outline

### 2019-06-12 draft.tex
finished a pragraph of proposed, a bit to go!

The fundamental barrier in writing down low-energy theories for strong correlated materials is the existence of non-perturbative interactions between various degrees of freedom - for example, lattice, spin and charge. Consider just the cuprate case for now, the two phases which are well understood contain a singular important degree of freedom in the low-energy space. Near zero doping the cuprates are charge-transfer insulators with an optical gap around 2 eV \textbf{ref}, indicating that the lowest energy excitations are primarily spin-like, allowing for a simple nearest-neighbor spin-only model like an AFM Heisenberg model to describe the low-energy physics quite accurately. In the heavily overdoped side, the lowest energy excitations look like well defined Fermi-liquid quasiparticles \textbf{ref}, indicating our lowest energy excitations are electron-like, and a simple non-interacting electron model (with renormalized masses, etc) will suffice. In the intermediate doping region where the low-energy excitations cannot be easily sectioned into spin-like or electron-like only, we see the emergence of exotic phases and where the most difficulties arise in writing down low-energy effective models. On top of that, this entire discussion did not include any lattice effects like structural phase transitions which many strongly-correlated materials have as well \textbf{ref}. This discussion can be extended to the many other strongly correlated electron materials presented above as well.

@@

We propose using a density matrix downfolding (DMD) procedure\textbf{ref} to develop low-energy effective theories of the 1-band nearest-neighbor Hubbard model on a square lattice as a stepping stone towards building effective theories for strongly correlated materials. The Hubbard model is a simple model with nearest-neighbor hopping and a local on-site Coulomb repulsion, parameterized by a single ratio U/t: 
\begin{equation}
H_\text{hub} = -t \sum_{\langle i,j \rangle,\sigma}( c_{i^\dagger,\sigma} c_{j,\sigma} + h.c.) + U \sum_i n_{i\uparrow} n_{i\downarrow}
\label{hubbard}.
\end{equation}
The solutions in 1-d are well understood via the Bethe ansatz \textbf{ref}, however the 2-d square lattice Hubbard model has no known exact solution.  In the very strong coupling limit $U/t \rightarrow \infty$ an emergent low-energy spin-space exists with an effective nearest-neighbor AFM Heisenberg Hamiltonian \textbf{ref}. In the region $U = 0$ a trivial non-interacting theory holds. Between these two limits, in particular the intermediate coupling region $1 < U < 7$, no well established effective theory exists. In this region there is a metal-insulator transition at $U_c/t \sim 4$ \textbf{ref}, and the pseudogap opens up at finite temperature around $U/t \sim 7$, leading to the development of a Mott gap for large $U/t > 11$ \textbf{ref}. In addition to the half-filling phase diagram, this model exhibits a plethora of interesting ground state properties when doped including stripe phases and d-wave pairing \textbf{ref}. The relative simplicity of the Hubbard model and the lack of a well-established low-energy theory in a broad intermediate coupling region lends the model as a useful playground for our new downfolding procedure in preparation for tackling realistic strongly correlated electron systems.

@@

Parameterizing our effective model requires constructing a basis on which to calculate our \textit{ab-initio} 1-/2-RDM elements and selecting a set of descriptors from which we would like to build our model.
We constructed two such bases: a localized intrinsic atomic orbital (IAO) basis and a molecular orbital (MO) basis.
The IAO basis was built on all Cu 3d, 4s and O 2p molecular orbitals from the $|\text{D}_i\rangle$ used to construct our sample states.
The MO basis consists of the Cu 3d, 4s and O 2p molecular orbitals from the lowest energy restricted open-shell Kohn-Sham (ROKS) solution with $x \leftrightarrow y$ spatial symmetry, a O 2p$_z \rightarrow$ O 2p$_\pi$ excitation.
As these two bases are used to describe the same space, they are related by an orthogonal transformation which will allow us to map between them if necessary.
In order to describe the low-energy degrees of freedom in the CuO molecule our model will require occupation energies and hybridizations between the Cu 3d, 4s and O 2p atomic orbitals which are most compactly represented in an MO basis.
The lack of doubly occupied Cu 4s states in the low-energy spectrum of CuO indicate the necessity of a Hubbard U term on the Cu 4s.
This is corroborated by the fact that the FN-DMC energy of the lowest-lying sampled state with a doubly occupied Cu 4s orbital is 7 eV above the lowest energy sampled state.
Further, we include a Hund's coupling between the Cu 4s and Cu 3d orbitals in our model based on the existence of a similar Hund's coupling 
for the copper atom.
These two interaction terms are best represented in the IAO basis as they are intuitively interactions between localized orbitals.
In addition to these important parameters, we will consider models which contain any number of hybridizations between the MOs to account for orbital relaxation between excited states in the CuO molecule.
In total our space of canditate models is then 
\begin{equation}
\begin{split}
\{\bar{n}_{4s}, \bar{n}_{p_\pi}, \bar{n}_{p_z}, \hat{n}_{4s\uparrow} \hat{n}_{4s\downarrow},\sum_{i \in \{xy, xz, ...\}}\vec{S}_{4s}\cdot \vec{S}_{d_i}\} \ + \\
\text{P}(\bar{c}_{d_\pi}^\dagger \bar{c}_{p_\pi} + h.c.,\ \bar{c}_{d_z^2}^\dagger \bar{c}_{p_z} + h.c.,\ \bar{c}_{4s}^\dagger \bar{c}_{p_z} + h.c.,\ \bar{c}_{d_z^2}^\dagger \bar{c}_{4s} + h.c.)
\end{split}
\label{eq:models}
\end{equation}
where P denotes the power set and operators are defined on the IAO basis unless denoted by a bar, in which case the operator is built on the MO basis. The coefficients for each term above will be denoted $\bar{\epsilon}_{4s},\ \bar{\epsilon}_{p_\pi},\ \bar{\epsilon}_{p_z},\ U_s,\ J_{sd},\ \bar{t}_\pi,\ \bar{t}_{dz},\ \bar{t}_{sz},\ \bar{t}_{ds}$. The operator $\bar{n}_{3d}$ is not included in any models as it is linearly dependent on all other occupation energies through the relationship $\bar{n}_{3d} + \bar{n}_{p_z} + \bar{n}_{p_\pi} + \bar{n}_{4s} = N_\text{elec} = \text{15}$. The symbol $\pi$ denotes a contraction over $x, y$ for p orbitals and $xz,\ yz$ for d orbitals. The symbol $\delta$, introduced later, denotes a contraction over $xy,\ x^2-y^2$ for d orbitals.

@@

>>>>

The fundamental barrier in writing down low-energy theories for strong correlated materials is the existence of non-perturbative interactions between various degrees of freedom - for example, lattice, spin and charge. Consider just the cuprate case for now, the two phases which are well understood contain a singular important degree of freedom in the low-energy space. Near zero doping the cuprates are charge-transfer insulators with an optical gap around 2 eV \textbf{ref}, indicating that the lowest energy excitations are primarily spin-like, allowing for a simple nearest-neighbor spin-only model like an AFM Heisenberg model to describe the low-energy physics quite accurately. In the heavily overdoped side, the lowest energy excitations look like well defined Fermi-liquid quasiparticles \textbf{ref}, indicating our lowest energy excitations are electron-like, and a simple non-interacting electron model (with renormalized masses, etc) will suffice. In the intermediate doping region where the low-energy excitations cannot be easily sectioned into spin-like or electron-like only, we see the emergence of exotic phases and where most difficulties arise in writing down low-energy effective models. On top of that, this entire discussion did not include any lattice effects like structural phase transitions which many strongly-correlated materials have as well \textbf{ref}.

@@

We propose using a density matrix downfolding (DMD) procedure\textbf{ref} to systematically develop low-energy effective theories of the 1-band nearest-neighbor Hubbard model on a square lattice as a stepping stone towards building effective theories for strongly correlated materials. The Hubbard model is a simple model with nearest-neighbor hopping and a local on-site Coulomb repulsion, parameterized by a single ratio U/t: 
\begin{equation}
H_\text{hub} = -t \sum_{\langle i,j \rangle,\sigma}( c_{i^\dagger,\sigma} c_{j,\sigma} + h.c.) + U \sum_i n_{i\uparrow} n_{i\downarrow}
\label{hubbard}.
\end{equation}
The solutions in 1-d are well understood via the Bethe ansatz \textbf{ref}, however the 2-d square lattice Hubbard model has no known exact solution.  In the very strong coupling limit $U/t \rightarrow \infty$ an emergent low-energy spin-space exists with an effective nearest-neighbor AFM Heisenberg Hamiltonian \textbf{ref}. In the region $U = 0$ a trivial non-interacting theory holds. Between these two limits, in particular the intermediate coupling region $1 < U < 7$, no well established effective theory exists. In this region there is a metal-insulator transition at $U_c/t \sim 4$ \textbf{ref}, and the pseudogap opens up at finite temperature around $U/t \sim 7$, leading to the development of a Mott gap for large $U/t > 11$ \textbf{ref}. In addition to the half-filling phase diagram, this model exhibits a plethora of interesting ground state properties when doped including stripe phases and d-wave pairing \textbf{ref}. The relative simplicity of the Hubbard model and the lack of a well-established low-energy theory in a broad intermediate coupling region lends the model as a useful playground for our new downfolding procedure in preparation for tackling realistic strongly correlated electron systems.

@@

Parameterizing our effective model requires constructing a basis on which to calculate our \textit{ab-initio} 1-/2-RDM elements and selecting a set of descriptors from which we would like to build our model.
We constructed two such bases: a localized intrinsic atomic orbital (IAO) basis and a molecular orbital (MO) basis.
The IAO basis was built on all Cu 3d, 4s and O 2p molecular orbitals from the $|\text{D}_i\rangle$ used to construct our sample states.
The MO basis consists of the Cu 3d, 4s and O 2p molecular orbitals from the lowest energy restricted open-shell Kohn-Sham (ROKS) solution with $x \leftrightarrow y$ spatial symmetry, a O 2p$_z \rightarrow$ O 2p$_\pi$ excitation.
We determine whether each basis is valid by looking at how much the trace of 1-RDM in FN-DMC on that basis differs from the expected value of 15 electrons.
We find that within 95\% confidence the trace is 14.7-14.8 (MO) and 14.6 - 14.7 (IAO), indicating that the major variation between states in $\mathcal{LE}$ can be described by either single particle basis.
As these two bases are used to describe the same space, they are related by a  transformation which will allow us to map between them if necessary.
As these two bases are used to describe the same space, they are related by an orthogonal transformation which will allow us to map between them if necessary.
In order to describe the low-energy degrees of freedom in the CuO molecule our model will require occupation energies and hybridizations between the Cu 3d, 4s and O 2p atomic orbitals which are most compactly represented in an MO basis.
The lack of doubly occupied Cu 4s states in the low-energy spectrum of CuO indicate the necessity of a Hubbard U term on the Cu 4s.
This is corroborated by the fact that the FN-DMC energy of the lowest-lying sampled state with a doubly occupied Cu 4s orbital is 7 eV above the lowest energy sampled state.
Further, we include a Hund's coupling between the Cu 4s and Cu 3d orbitals in our model based on the existence of a similar Hund's coupling 
for the copper atom.
These two interaction terms are best represented in the IAO basis as they are intuitively interactions between localized orbitals.
In addition to these important parameters, we will consider models which contain any number of hybridizations between the MOs to account for orbital relaxation between excited states in the CuO molecule.
In total our space of canditate models is then 
\begin{equation}
\begin{split}
\{\bar{n}_{4s}, \bar{n}_{p_\pi}, \bar{n}_{p_z}, \hat{n}_{4s\uparrow} \hat{n}_{4s\downarrow},\sum_{i \in \{xy, xz, ...\}}\vec{S}_{4s}\cdot \vec{S}_{d_i}\} \ + \\
\text{P}(\bar{c}_{d_\pi}^\dagger \bar{c}_{p_\pi} + h.c.,\ \bar{c}_{d_z^2}^\dagger \bar{c}_{p_z} + h.c.,\ \bar{c}_{4s}^\dagger \bar{c}_{p_z} + h.c.,\ \bar{c}_{d_z^2}^\dagger \bar{c}_{4s} + h.c.)
\end{split}
\label{eq:models}
\end{equation}
where P denotes the power set and operators are defined on the IAO basis unless denoted by a bar, in which case the operator is built on the MO basis. The coefficients for each term above will be denoted $\bar{\epsilon}_{4s},\ \bar{\epsilon}_{p_\pi},\ \bar{\epsilon}_{p_z},\ U_s,\ J_{sd},\ \bar{t}_\pi,\ \bar{t}_{dz},\ \bar{t}_{sz},\ \bar{t}_{ds}$. The operator $\bar{n}_{3d}$ is not included in any models as it is linearly dependent on all other occupation energies through the relationship $\bar{n}_{3d} + \bar{n}_{p_z} + \bar{n}_{p_\pi} + \bar{n}_{4s} = N_\text{elec} = \text{15}$. The symbol $\pi$ denotes a contraction over $x, y$ for p orbitals and $xz,\ yz$ for d orbitals. The symbol $\delta$, introduced later, denotes a contraction over $xy,\ x^2-y^2$ for d orbitals.

@@

\section{Proposed work}
We want to develop low-energy effective theories for the
half-filled Hubbard model over the space U: 0 - 12, 1/k$_B$T: 0 - 3 using the DMD method, particularly focusing on the medium coupling region of U: 2 - 6 and 1/k$_B$T $\sim$ 1 where perturbative effective theories break down and a metal-insulator transition exists.
Similar to the case of the \textit{ab-initio} downfolding, we will need to construct low-energy spaces, sample the constructed spaces, build effective theories and respective bases, and fit our models.
Unlike \textit{ab-initio} downfolding, the Hubbard model downfolding has significantly fewer degrees of freedom, and therefore the DMD procedure can be done in a much more controlled fashion.
We will begin from a starting point of highest control, namely for system sizes small enough where we can exactly diagonalize the Hubbard model.
In the situation where we have full access to the eigenstates and spectra of a system, selecting $\mathcal{LE}$ can be done exactly as the span of some chosen $N$ lowest energy eigenstates. 
We can further guarantee that our sample set fully encapsulates all low-energy degrees of freedom by simply sampling states parameterized as $|\Psi\rangle = \sum_{j=1}^N c_j |\Phi_j\rangle$ where $|\Phi_j\rangle$ are eigenstates of the Hubbard Hamiltonian.
This can be seen in contrast to the CuO fitting where $\mathcal{LE}$ was defined implicitly through a number occupation descriptor and the sampling of $\mathcal{LE}$ was incomplete leading to intruder states in our initial effective theories.
Basis construction for the Hubbard Hamiltonian is also explicitly clear as any possible single particle basis we can use is just an orthogonal transformation of the basis the Hubbard model is defined on, and is guaranteed to be a complete single particle basis.
Again, this can be compared to the case of an \textit{ab-initio} system like the CuO molecule where bases are only approximately complete due to the huge number of degrees of freedom in the system.
In order to include temperature into the mix we can work with the full model Hilbert space such that $\mathcal{LE} = \mathcal{H}$, but generate samples within our space using a Monte Carlo sampling method according to a Boltzmann probability distribution with variable parameter $\beta = 1/k_BT$.
This way, our sample space will still be able to probe every excitation within a certain temperature range, the samples will be biased towards lower energy states according to the Boltzmann weight, and hence the fit model will also be biased towards describing more accurately the energy functional on the states with highest weight.
The Monte Carlo sampling may need to be altered further in order to deal with entropic effects when the density of states is very low in an energy range of interest and relatively high elsewhere.

Show in \ref{fig:Hubbard} is ... \textbf{SHOW OUR PRELIMINARY RESULTS}

In order to make any statements about effective theories for the Hubbard model we will have to work with larger system sizes and approximate methods like lattice quantum Monte Carlo to generate low-energy states for the DMD procedure. \textbf{TALK ABOUT FINITE SIZE SCALING AND APPROXIMATE METHODS (liken to FN-DMC for ab initio)}

A farther reaching goal would be to develop effective theories in the U-T-$\delta$ (doping) space for the Hubbard model, where more complex behavior similar to the cuprates superconductors have been found.
